---
title: '20240920'
author: "Data350"
date: "2024-09-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(socviz)
library(nycflights13)
library(gapminder)
```


# Kernel Density Estimates

Let's play around with geom_histogram(), geom_density(), stat_bin(), and stat_density().

1.    Consider the diamonds dataset. The *price* can be viewed as a continuous variable, so density is the key. Create a character list clar of the worst, medium, and the best clarity grade. Plot the price density of each grade. Try scale_colour_viridis_d()

```{r}
clar <- c("I1", "VS1", "IF")
ggplot(
  data = subset(diamonds, clarity %in% clar),
  mapping = aes(x = log(price))
  ) + geom_density(
    aes(fill = clarity),
    alpha = 0.4
    )
```



2.    Plot the density along with a histogram. How would you work around it? Try using "y=after_stat(density)" in your histogram aesthetics. What does that do? Tune the color and transparency to make the pictures nicer. You can use "scale_fill_manual()" and "scale_color_manual()" to manually change the colors. Add labels and captions. 

```{r}
ggplot(
  data = subset(diamonds, clarity %in% clar),
  mapping = aes(x = log(price))
  ) + 
geom_density(
    aes(color = clarity),
    alpha = 1
    ) +
geom_histogram(
  bins = 40,
  aes(
    y=after_stat(density), 
    fill = clarity), 
  position = 'identity',
  alpha = 0.6
  ) + 
scale_color_manual(
  values = c("red","blue","green")) + 
scale_fill_manual(
  values = c("pink","lightblue","lightgreen")) + 
  labs(
    x = "Log Price", 
    y = "Density", 
    caption= "Blah", 
    title = "Blah Blah")
```



# More Multi-facted plots

```{r}
library(gridExtra)
```


Continue with the diamonds dataset. Zoom in on the price of clarity grade "IF"

```{r}
diamonds2 <- diamonds |> filter(clarity == "IF")
```


1.    Create four plots, g1 through g4. Each plot counts the density of price. In the arguments for stat_density(), use "gaussian" (default), "epanechnikov", "triangular", and "cosine". See ?density() for other kernel options. 

```{r}
g1 <- ggplot(diamonds2, aes(x = log(price))) +
  stat_bin(aes(x = log(price), y = after_stat(density))) + stat_density(kernel = "gaussian", color = "red", fill = NA)

g2 <- ggplot(diamonds2, aes(x = log(price))) +
  stat_bin(aes(x = log(price), y = after_stat(density))) + stat_density(kernel = "epanechnikov", color = "red", fill = NA)

g3 <- ggplot(diamonds2, aes(x = log(price))) +
  stat_bin(aes(x = log(price), y = after_stat(density))) + stat_density(kernel = "triangular", color = "red", fill = NA)

g4 <- ggplot(diamonds2, aes(x = log(price))) +
  stat_bin(aes(x = log(price), y = after_stat(density))) + stat_density(kernel = "cosine", color = "red", fill = NA)
```

2.    Use "grid_arrange(g1,g2,g3,g4, nrow = 2)" and see what you get. 

```{r}
grid.arrange(g1, g2, g3, g4, nrow = 2)
```
3.    Tune everything to make things look nicer.

4.    A different method: borrow the code from part 2 of the last section and add "facet_wrap(~color)".

```{r}
ggplot(
  data = subset(diamonds, clarity %in% clar),
  mapping = aes(x = log(price))
  ) + 
geom_density(
    aes(color = clarity),
    alpha = 1
    ) +
geom_histogram(
  bins = 20,
  aes(
    y=after_stat(density), 
    fill = clarity), 
  position = 'identity',
  alpha = 0.6
  ) + 
  facet_wrap(~color) +
scale_color_manual(
  values = c("red","blue","green")) + 
scale_fill_manual(
  values = c("pink","lightblue","lightgreen")) +
  labs(
    x = "Log Price", 
    y = "Density", 
    caption= "Blah", 
    title = "Blah Blah")
```



# 2D Kernel Density Estimates

```{r}
library(hexbin)
```



```{r}
set.seed(2024)
ind <- sample(53940, 2000)
diamonds2 <- diamonds[ind,]
```


1.    In the diamonds dataset, Create a subsample of size 2000. Plot the price (y) against carat(x). Tune the paramters to make things look nicer. 

```{r}
g <- ggplot(diamonds2,
       aes(x = carat, y = log(price))
       )
g + geom_point(size = 0.5)
```



2.    Read and try "geom_bin2d()", "geom_hex()", and "geom_density_2d()"

```{r}
g +  
  geom_hex(bins = 25) +
  geom_point(alpha = 0.5, shape = "o", size = 0.5, color = "black") +
  scale_fill_gradient(low="lightgreen",high="red",trans="log10")
```

```{r}
g +  
  geom_density2d(color = "red") +
  geom_point(alpha = 0.5, shape = "o", size = 0.5, color = "black") +
  geom_hex(bins = 25, alpha = 0.5) + scale_fill_gradient(low="lightgreen",high="blue",trans="log10")
```



# Visualizing confidence intervals

## Exercise 1. More *geom_smooth()*

1.  In the *gapminder* dataset, only consider the data in the years 1952, 1972, and 1992.

2.  Create a point plot of life expectancy vs GDP per capita, and *Create a regression curve for each year*.

-   Use color to different the years, and use size to describe population size.

-   Play around with different regression models to find the best one (to tell your story).

-   You may want to rescale the axis to make things look nicer. Type in "scale" in the Help window and take a look.

4.  Fix the legends, pick themes, etc.

5.  Write everything using the pipe (if you haven't already).

```{r}
  gapminder |>
  filter(year %in% c(1952, 1972, 1992)) |>
  mutate(year = as.factor(year)) |> 
  ggplot(
    mapping = aes(x = gdpPercap, y = lifeExp, color = year)
  ) + 
  geom_point(aes(size = pop)) +
  scale_x_log10(labels = scales::dollar) + 
  geom_smooth(method = "glm")

```

## Excercise 2: Confidence Intervals with bar plot

Let's re-create the plots in the slides.

1.  In the gapminder dataset, compute the mean and the confidence intervals ($\mathsf{Mean }\pm 2\cdot \mathsf{StdDev}$ ), grouped by continents and years. Store it in a new dataframe called gapminder2.

```{r}
gapminder2 <- gapminder |> 
  group_by(continent, year) |>
  summarize(
    ave_life = mean(lifeExp),
    UCI_sd = ave_life + 2*sd(lifeExp),
    LCI_sd = ave_life - 2*sd(lifeExp),
    UCI_se = ave_life + 2*sd(lifeExp)/sqrt(length(lifeExp)),
    LCI_se = ave_life - 2*sd(lifeExp)/sqrt(length(lifeExp))
  )
```


2.  Use geom_bar() to create a bar plot for the years 1957, 1982, 2002. For each year, there should be five bars stacked next to each other, each representing a continent.

```{r}
p <- ggplot(
  data = subset(gapminder2, year %in% c(1957, 1982, 2002)),
  mapping = aes(x=year, y = ave_life)
  ) + 
  geom_bar(
    mapping = aes(fill = continent),
    stat = 'identity',
    position = position_dodge(), # the bars are stacked next to each other
  )
p
```


3.  Add the error bars using geom_`errorbar()

4.  Add the labels and fine-tune the aesthetics to make things look nicer if you like.

5.  Create a new dataframe gapminder3, where you calculate the average life expectancy by year only, and you also compute two intervals, one using standard error, the other one using standard deviation.

6.  Create a line plot of average life expectancy vs years with standard deviation error bars.

7.  Create a separate line plot with standard error.

## Exercise 3: Bar plot for flights

1.  Let's create a **bar plot** counting the monthly average departure delay for each carrier in the *flights* dataset.

2.  Visualize the confidence intervals
