---
title: 'Multi-panel plots'
author: "Data350"
date: "2024-10-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(socviz)
library(gapminder)
library(ggplot2)
```

# facet_wrap()

1.  Recall that the following block of codes produce a bar chart of political inclination in the US. Modify the code to produce a 2-by-2 plot and a 1-by-1 plot according to big region. Try wrapping around different (categorical) variables.

```{r}
p <- ggplot(
  data = subset(gss_sm, partyid !="NA"),
  mappping = aes(
    x = partyid,
  )
)
p + geom_bar(
  position = "dodge",
  mapping = aes(
    x = partyid,
    y = ..prop..,
    group = bigregion
  )
) + coord_flip() + facet_wrap(~happy, ncol = 1)

```

2.  In the diamonds dataset, create a $1\times 5$ grid of boxplots of *clarity vs price*, each one corresponding to the quality of the cut. Scale the y-axis by $\log_{10}$.

```{r}
ggplot(
  data = diamonds,
  mapping = aes(x = clarity, y = price)
) + 
  geom_boxplot(aes(color = cut)) + 
  facet_wrap(~cut, nrow = 1) + 
  scale_y_log10(labels = scales::dollar)
```

# facet_grid()

The formula:

-   "facet_grid(var1 ~ var2)" spread var1 across the rows and var2 across the columns
-   "facet_grid(. ~ var)" spreads var across the columns. "facet_grid(var ~ .)" spreads var across rows.

1.    In the mpg dataset, create a suitable grid of point plots of *engine displacement vs highway mileage*, and each point plot corresponds to a certain vehicle class and drive mode. Use color to differentiate the years made. 

```{r}
ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) + 
  geom_point(aes(color = as.factor(year)), size = 0.5) + 
  facet_grid(class ~ drv)
```

2.    Insert the labels by putting "labeller = labeller(drv = drv_labs, class = class_labs)" into facet_grid()

```{r}
library(scales)
drv_labs <- c("4" = "4 wheel drive", "f" = "Front wheel drive", "r" = "Rear wheel drive")
class_labs <- c("2seater" = "Two seater", "compact" = "Compact, my money don't jiggle jiggle", "midsize" = "Mid-size", "minivan" = "Minivan", "pickup" = "Pickup Truck", "subcompact" = "Subcompact", "suv" = "SUV")
ggplot(
  data = mpg,
  mapping = aes(x = hwy, y = displ)
) + 
  geom_point(aes(color = as.factor(year)), size = 0.5) + 
  facet_grid(
    drv ~ class,
    labeller = labeller(drv = as_labeller(drv_labs), 
                        class = as_labeller(class_labs, label_wrap_gen(width = 10)) # 10 here is the character limit
                        )
    ) 

```
In the example above, the "label_wrap_gen()" function helps you wrap long texts according to the specified widths. 



In facet_wrap() and facet_grid(), can control if
position scales are the same in all panels (fixed) or allowed to
vary between panels (free) with the scales parameter:

-   scales = "fixed” : x and y scales fixed
-   scales = "free_x” : x scale is free, y scale is fixed.
-   scales = "free_y” : y scale is free, x scale is fixed.
-   scales = "free” : x and y scales free.

3.    Consider the economics_long dataset and create a grid of plots for each of the "variable" (dimension of your own choosing) of *date vs value* (so you get some kind of time-series). Which parameter should you free up? The title_labs is for your labeling convenience. 

```{r}
title_labs <- c(
  pce = "Personal Consumption Expenditures (in $B)", 
  pop = "Total population in thousands", 
  psavert = "Personal savings rate", 
  uempmed = "Median duration of unemployment, in weeks", 
  unemploy = "Number of unemployed, in thousands"
)

ggplot(
  data = economics_long,
  mapping = aes(x = date, y = value)
) + 
  geom_line() + 
  facet_wrap(
    ~ variable,
    labeller = labeller(variable = title_labs),
    nrow = 2,
    scales = "free_y", # free up y variable to make trends more transparent,
  ) + 
  theme(axis.text.x = element_text(angle = 90))
```


4.    Consider joining the dataset oecd_le and gapminder by doing the following:

```{r}
library(tidyverse)
oecd2 <- left_join(oecd_le, gapminder, by = "country")
```

Create a point plot where:

-   each row is corresponds to a country, and the x-axis corresponds to the life expectancy
-   all the countries in the same continent are grouped together
-   set the scales and the space to the appropriate parameter (look up ?facet_grid())
-   Anything to fix?




5.    A time-series plot of flight delay for each airline using "stat_summary()". 

```{r}
library(nycflights13)
flights$date <- ISOdate(flights$year, flights$month, flights$day)
ggplot() + 
  stat_summary(
    fun = "mean",
    geom = "line",
    na.rm = T,
    col = "blue"
  )
```

Sort the airline by descending alphabetical order. Compare the average delay of the first five airlines vs the rest throughout a year. 

```{r}

```




# Manually arranging plot

```{r}
library(patchwork)
```


Create 4 plots p1 -- p4 for the mpg dataset:

1.  p1 is a scatter plot with engine displacement vs highway
2.  p2 is a bar chart with year vs drv
3.  p3 is a multipanel density plot of highway mpg according to drv
4.  p4 is given below:

```{r}
p1 <- ggplot(mpg) + geom_point(aes(x = displ, y = hwy))
p1
```

```{r}
p2 <- ggplot(mpg) + geom_bar(aes(
  x = as.factor(year),
  fill = drv
), position = "dodge") + labs(x = "year")

p2
```

```{r}
p3 <- ggplot(mpg) + geom_density(
  aes(x = hwy, fill = drv), color = NA
) + facet_wrap(~drv, ncol = 1)
p3
```


```{r}
p4 <- ggplot(mpg) + 
  stat_summary(
    aes(x = drv, y = hwy, fill = drv),
    geom= "col",
    fun.data = mean_se
    ) + 
  stat_summary(
    aes(x = drv, y = hwy),
    geom = "errorbar",
    fun.data = mean_se,
    width = 0.3
  )
p4
```

Try the following

```{r}
p1 + p2 + p3 + p4
```

-   You can use plot_layout() to manually arrange the plots.

```{r}
p1 + p2 + p3 + plot_layout(ncol = 2)
```

-   You can use | or / to force a single row or a single column, respectively:

```{r}
(p1 | p2) / (p3|p4)
```

-   You can use more complex layout:

```{r}
p1 + p2 + p3 + p4 + plot_layout(widths = c(5,2))
```
or design your own. 
```{r}
layout <- "
AABBB
C#BBB
C##DD
"
p1 + p2 + p3 + p4 + 
  plot_layout(
  design = layout,
  guides = "collect"
  )

```

```{r}
layout <- c(
  area(t = 2, l = 1, b = 5, r = 4),
  area(t = 1, l = 3, b = 3, r = 5)
)
p1 + p2 +
  plot_layout(design = layout)
```


-   You can use guides = "collect" in plot_layout() to remove redundant legends:

```{r}
p1 + p2 + p3 + plot_layout(ncol = 2, guides = "collect")

# p1 + p2 + p3 + guide_area() + plot_layout(ncol = 2, guides = "collect")
```

-   You can also join the plots (just like in powerpoint) to modify them collectively:

```{r}
p12 <- p1 + p2
```

To update each individual component:
```{r}
p12[[2]] <- p12[[2]] + theme_void()
```

You can use "&" to modify all plots

```{r}
# p1 + p4 & theme_void()

p_y_scale <- p1 + p4 & scale_y_continuous(limits = c(0,50))
p_y_scale
```

To summarize: patchwork shows mutiple plots side-by-side to tell a compelling story. After assemply, plots become a single object and anotation works on the whole figure:

```{r}
p34 <- p3 + p4 + plot_annotation(
  title= " A closer look at the effect of drve train in cars",
  caption = "Source: mpg dataset in ggplot2"
)
p34
```

-   Take a closer look at plot_annotation().


# Plots within plots

The star here is the inset_element() function. 

```{r}
p1 + inset_element(
  p2, 
  left = 0.5,
  bottom = 0.4,
  right = 0.9,
  top = 0.95
)
```


# Pairs Plots

“Pairs plots” are a convenient way to plot the relationships between multiple data dimensions.

-   Think of them as multidimensional scatterplots
-   Base *R* produces OK pairs plots (quick and dirty but...)
-   Shine with **ggplot2**


## Fisher's or Anderson's Iris data using Base R

From R: “This famous (Fisher’s or Anderson’s) iris data set gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris setosa, versicolor, and virginica.”

```{r}
X <- iris3
```

Take a look at iris3. What's its dimension. How to interpret read the entries?

1.  Suppose you want to create a dataframe/tibble called X_setosa that contains all the observations about the Setosa sepcies. What would you do?

```{r}
X_setosa <- X[,,1] #selecting every row and every column from the first slice
```


2.    How would you interpret the output?

```{r}
pairs(X_setosa)
```


3.    Recall rbind(). Create a dataframe called X_iris3 that includes all of the sepal and petal information for all species. Create a pairs plot. What do you notice about the (i,j)-th and the (j,i)-th grid?

```{r}
X_iris3 <- rbind(X[,,1], X[,,2], X[,,3])
pairs(X_iris3, 
      lower.panel = panel.smooth,
      upper.panel = NULL,
      lwd = 3
      )
```

4.    That means you have some extra room to insert more information. Try each of the following lines. 

```{r}
pairs(X_iris3, lower.panel = panel.smooth, lwd = 2, color = "blue")
pairs(X_iris3, lower.panel = panel.smooth, lwd = 2, color = "blue", upper.panel = NULL, main = "Iris Data")
pairs(X_iris3, col=c(rep(1,50),rep(2,50),rep(4,50)), main = "Iris Data")
```

```{r}
pairs(X_iris3, 
      col=c(rep(2,50),rep(3,50),rep(4,50)), 
      lower.panel = panel.smooth, lwd = 2,
      upper.panel = NULL,
      main = "Iris Data")
```

## Better plots using *GGally* package

```{r}
library(GGally)
```


1.    First we need to create a dataframe. Start with the following. What's missing after running the following line of code? How to fix it?

```{r}
df <- data.frame(X_iris3) # the species information is missing
df <- df |> mutate(
  species = c(
    rep("Setosa",50),
    rep("Versicolor",50),
    rep("Virginica" ,50)
  )
)
```


2.    Try the following.

```{r}
ggpairs(df)
```

3.    We can make things prettier. Modify the following block of code so that color corresponds to the species, tune the parameters (the syntax for ggpairs is the same as ggplot for all the basic layers).

```{r}
ggpairs(df, 
        columns = 1:4, # ignore the species column
        aes(color = species),
        upper = list(continuous = wrap("cor", size = 2.5)), # upper right corner
        lower = list(continuous = wrap("points", size = 1.0)), # lower left corner
        diag = list(continuous = wrap("densityDiag", alpha = 0.5)), # diagonal
        title = "Pairs Plot of Iris Data",
        )
```


3.    Pairs-plot columns (1,2,3,5) and see what you get:

```{r}

```


