---
title: "Geospatial Maps"
author: "Data350"
date: "2024-10-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Goal: produce plots of geospatial data**

# Getting started

```{r}
library(maps)
library(ggplot2)
library(tidyverse)
```

-   The basic commands:

    -   For US domestics

```{r}
maps::map("state", fill = T, col = palette())
```

```         
-   For any region in the world (many of them do not have internal state boundaries)
```

```{r}
maps::map("world", region = "Spain")
# Canada, Australia, Brazil, Italy, Japan, France, UK
```
```{r}
maps::map("world", regions = "canada")
```


```{r}
maps::map("world", regions = "")
```

```         
-   You can change colors by specifying *col = ...*
```

```{r}
maps::map("state", fill = TRUE, col = terrain.colors(49))
# the number here tells you how many colors in the palette to use
# maps::map("state", fill = TRUE, col = rainbow(49))
maps::map("state", fill = TRUE, col = heat.colors(49, alpha = 0.7))
```

# A local example

```{r}
maps::map(
  "county", 
  "maryland",
  fill = TRUE,
  col = heat.colors(10, alpha = 0.5)
)
```
*Exercise: change the Maryland plot to Maryland colors*

```{r}

```

The following code will extract the information about Maryland counties

```{r}
head(map_data("county", "maryland"))
```

-   Create a new dataframe called md_counties by

    -   selecting the columns containing longitude, latitude, group, and subregion and
    -   changing the name of "long" to "lon" and "subregion" to "id".

    ```{r}
md_counties <- map_data("county", "maryland") |>
      select(
        lon = long,
        lat,
        group,
        id = subregion
      )
str(md_counties)
    ```

    -   Create a point plot of the dataset to see what each coordinates represent. Make sure to add a layer *coord_quickmap()* for better spatial geometry.
    
```{r}
ggplot(md_counties,
       aes(x = lon, y = lat)
       ) + 
  geom_point(size = 0.2, aes(color = id)) + 
  coord_quickmap() + # line up longitude with latitude
  labs(x = "Longitude", y = "Latitude", color = "County")
```


# Polygons and Simple Features

The key to complete the map is the *geom_polygon()* function.

-   allows us to draw polygons around objects by group

-   connect the boundary points of a region using lines

```{r}
ggplot(md_counties, 
       aes(lon, lat, group = group)
       ) + 
  geom_polygon(color = "white", aes(fill = id)) +
  coord_quickmap() +
  theme_dark() 
```

# Medical cases in North Carolina

What if the data is not lat/lon for boundaries?

-   often map data is vectorized and encoded using the “simple features” standard of the **Open Geospatial Consortium**
-   the *sf* (simple feature) package allows us to handle this

```{r}
library(sf)
```

```{r}
nc <- sf::st_read(
  system.file("shape/nc.shp", package = "sf"),
  quiet = TRUE
  )
```

Take a loot at "nc" and plot it by adding two layers *geom_sf()* and *coord_sf()*

```{r}
ggplot(nc) + 
  geom_sf() + # plots the polygon
  coord_sf() # line up the longitudes```
```


The variable "SID74" is the number of sudden infant death in the period 1974-78. You can plot this by county by specifying the *fill* aesthetics in *geom_sf()*. Adjust the colors using *scale_fill_gradient()*

```{r}
ggplot(nc) + 
  geom_sf(aes(fill = SID74)) + 
  coord_sf() + 
  scale_fill_gradient(
    low = "white",
    high = "red"
  ) + 
  labs(
    title = " Sudden Infant Death Syndrom by County in North Carolina 1974 ",
    fill = "Sudden Infant Deaths\n in North Carolina 1974",
    caption = "Source: nc.shp dataset"
    )
```
Change it into a percentage.

```{r}
library(scales)
ggplot(nc) + 
  geom_sf(aes(fill = (SID74/BIR74))) + 
  coord_sf() + 
  scale_fill_gradient(
    low = "white",
    high = "red",
    labels = label_percent()
  ) + 
  labs(
    title = "SID case density in North Carolina by county, 1974-1978",
    caption = "Source: nc dataset from sf",
    fill = "%(# cases)/(# new births)",
    ) + 
  theme(plot.title = element_text(hjust = 0.5)) # adjust the position of the title
```
Make a comparison plot between 1974-1978 and 1979-1983:

```{r}
range_values <- range(
  c(
    nc$SID74 / nc$BIR74, 
    nc$SID79 / nc$BIR79
    ), 
  na.rm = TRUE) # gives me the maximum range of the scale

p74 <- ggplot(nc) + 
  geom_sf(aes(fill = (SID74/BIR74))) + 
  coord_sf() + 
  scale_fill_gradient(
    low = "white",
    high = "red",
    labels = label_percent(),
    limits = range_values # specify the limits of the scale
  )

p79 <- ggplot(nc) + 
  geom_sf(aes(fill = (SID79/BIR79))) + 
  coord_sf() + 
  scale_fill_gradient(
    low = "white",
    high = "red",
    labels = label_percent(),
    limits = range_values # specify the limits of the scale
  )
```

Use patchwork to put them together

```{r}
library(patchwork)

p74 / p79
```

Do a difference plot

```{r}
ggplot(nc) + 
  geom_sf(aes(fill = (SID79/BIR79)-(SID74/BIR74))) + 
  coord_sf() + 
  scale_fill_gradient2(
    low = "green",
    mid = "white",
    high = "red",
    labels = label_percent(),
  )
```

Add some text to the plot using "ggrepel"

```{r}
library(ggrepel)
```


Import another dataset from maps:

```{r}
library(tidyverse)

# select the top five populous city in NC
nc_cities5 <- maps::us.cities |> filter(country.etc == "NC") |> arrange(desc(pop)) |> slice_head(n=5)

nc_cities5 
```

Overlap the SID plot with texts

```{r}
options(repr.plot.width = 25, repr.plot.height = 14)

ggplot() + 
  geom_sf(data = nc,
    aes(fill = (SID79/BIR79)-(SID74/BIR74))) + 
  coord_sf() + 
  scale_fill_gradient2(
    low = "green",
    mid = "white",
    high = "red",
    labels = label_percent(),
  ) + 
  geom_point(
    data = nc_cities5,
    mapping = aes(x = long, y = lat),
    size = 2,
    color = "purple"
  ) + 
  geom_text_repel(
    data = nc_cities5,
    mapping = aes(x = long, y = lat, label = name),
    color = "black", 
    size = 2.5
  )
```



# Covid Dataset by counties in MD

Extract the geospatial data of MD:

```{r}
md_counties <- map_data("county", "maryland")
head(md_counties)
```

Change subregion to "id", and only want columns of coordinates and group and id.

```{r}
md_counties <- md_counties |> 
  select(
    lon = long,
    lat, 
    group, 
    id = subregion
  )
head(md_counties)
```

After downloading the dataset from <https://data.imap.maryland.gov/datasets/maryland>::mdcovid19-casesbycounty/about

```{r}
mdcovid <- read.csv("MDCOVID19_CasesByCounty.csv")
```

Take a look:

```{r}
head(mdcovid)
```

Say we are interested in the total number of cases throughout the period of the survey, we need to take column sum:

```{r}
county_sum <- colSums(mdcovid[, -c(1,2)]) # not summing the first (ID) or the second (timestamp) columns
county_sum
```

```{r}
names(county_sum)
```

Need to make some kind of alignment between the two datasets. Let's align with md_counties, i.e., remove "\_" from mdcovid and make it lower-case:

```{r}
# change every character in names(county_sum) to lower-case
names(county_sum) <- tolower(names(county_sum)) 

# replace "_" by an empty space
names(county_sum) <- sub("_", " ", names(county_sum)) 


names(county_sum)
```

```{r}
county_sum <- filter(data.frame(county_sum), county_sum!= 0)
county_sum
```

Join the dataset

1.  Create a new column called "id" to match with md_counties

```{r}
county_sum$id <- rownames(county_sum)
county_sum
```

2.  spread the county number according to id using **join**

```{r}
md_counties <- left_join(
  md_counties,
  county_sum,
  by = "id"
)
# md_counties$case_num <- md_counties$county_sum
head(md_counties)
```

Plot it out

```{r}
library(scales)

ggplot(md_counties,
       aes(lon, lat, group = group)
       ) + 
  geom_polygon(aes(fill = log(county_sum)), # using a log scale here
               color = "black", linewidth = 0.2) + 
  coord_quickmap() + 
  scale_fill_gradient2(
    low = "green", 
    mid = "white", 
    high = "red",
    midpoint = log(median(md_counties$county_sum)), # re-center the scale
    limits = log(c(min(md_counties$county_sum),  max(md_counties$county_sum))), # change the range
    labels = math_format(10^.x) # make sure the legend is also in log scale
    ) + 
  labs(
    fill = "Case Number",
    title = "Covid cases by MD counties\n 03/15/2020 -- 10/22/2024",
    caption = "Source: https://data.imap.maryland.gov"
  )
```

