---
title: "Data Visualization in R"
subtitle: "Correlation and Pairs Plot"
author: "Fushuai Jiang"
date: "`r Sys.Date()`"
output: beamer_presentation

urlcolor: blue

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  message = FALSE)
knitr::opts_chunk$set(fig.align='center', out.width = "75%")
library(ggplot2)
library(tidyverse)
library(patchwork)
library(ggcorrplot)
library(corrplot)
library(GGally)
```




# Heatmap basic


## Heat map:




# Hierarchical Clustering

## Example: `iris`

First we perform PCA to separate the data

```{r, out.width = "60%"}
set.seed(2025)
pc <- prcomp(iris[,1:4],retx=T)
ind <- sample(150,30)
pc <- data.frame(pc$x[ind,1:2])
pc$Species <- iris$Species[ind]
ggplot(pc,aes(x=PC1,y=PC2)) + geom_point(aes(color=Species),size=4) +
  ylim(-3.5,3.5) + xlim(-3.5,3.5)
```

## Example: `iris`

```{r}
#install.packages("pheatmap")
df <- dist(pc[,1:2])
col_annotation <- data.frame(pc$Species)
colnames(col_annotation) <- c("Species")
pheatmap::pheatmap(df,annotation_col = col_annotation)
```

## Example: `iris`

```{r, echo=F, out.width="100%"}
#install.packages("pheatmap")
df <- dist(pc[,1:2])
col_annotation <- data.frame(pc$Species)
colnames(col_annotation) <- c("Species")
pheatmap::pheatmap(df,annotation_col = col_annotation)
```

## The linkage matters (default: `complete`)

```{r}
df<-dist(pc[,1:2])
plot(hclust(df))
```


# Signal Processing



## Spectral Density & Periodogram (Intuition)

**Spectral Analysis?**
  - Time series can have hidden cycles (daily, weekly, yearly patterns).
  - Instead of focusing only on *time domain* (plot vs. time), we also look in the *frequency domain*.

**Spectral Density**
  - Tells us how much of the series’ variability comes from cycles of different frequencies.
  - Low frequencies = slow changes, long-term trends.  
  - High frequencies = rapid changes, short-term noise.

**Periodogram**
  - A way to estimate spectral density from observed data.
  - Peaks in the periodogram show which cycles (frequencies) are strongest.
  
  
  
## The math
  
- **Discrete Fourier Transform (DFT):**  
$$
d_n(\omega_k) = \frac{1}{\sqrt{n}} \sum_{t=0}^{n-1} X_t \, e^{-i \omega_k t}, 
\quad \omega_k = \frac{2\pi k}{n}, \; k=0,\dots,n-1
$$

- **Periodogram (spectral density estimator):**  
$$
I_n(\omega_k) = \frac{1}{2\pi n} \left| \sum_{t=0}^{n-1} X_t e^{-i \omega_k t} \right|^2
= \frac{1}{2\pi} |d_n(\omega_k)|^2
$$

- **Properties:**  
  - Peaks in \(I_n(\omega)\) indicate strong periodic components.  
  - High variance → smoothing or averaging recommended.
  

  
  


## Spectral Density Estimation via `spectrum`

```{r}
x <- economics_long |> dplyr::filter(variable == "pce") |> pull(value)
spe <- spectrum(x)
```

## Periodogram

```{r}
str(spe)
```





## Spectrogram

```{r}
library(signal)
x <- economics_long |> dplyr::filter(variable == "pce") |> pull(value)
sp <- specgram(x, n = 64)
sp
```

## Spectrogram

```{r}
jet <- grDevices::colorRampPalette(c("#00007F", "blue", "#007FFF", "cyan","#7FFF7F", "yellow", "#FF7F00", "red", "#7F0000"))
plot(sp, col = jet(20))
```

