---
title: "Measles"
author: "Fushuai Jiang"
date: "2025-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(ggplot2)
library(tidyverse)
library(sf)
library(readr)
library(leaflet)
library(gridExtra)
library(patchwork)
```



## Loading data
```{r}

path = "C:/Users/black/Documents/GitHub/dataviz/Measles"
setwd(path)
remove(path)

measles_state <- read_csv('measles_by_state.csv') 
measles_year <- read_csv('measles_by_year.csv')
measles_week <- read_csv('measles_by_week.csv')
vaccine_state <- read_csv('vaccine_by_state.csv')
```

## Cleaning the dataset



- The "measles_by_year" dataset has a column named "filered_by..." and it is not relevant to us

- We want to add a single "week" column to "measles_by_week" for plotting purpose, and we set it to be the start of the week. We also want to change the variable type from chr to date using the function "as.Date()"

- The "measles_by_state" dataset has a column named "cases_range", which are stored as characters. We would like to create a new column taking a numeric value representing the range, say the average or the max. 

- The percentage in vaccine dataset is stored as character so we need to change it into numerics


```{r}
# remove the "filtered by..." columns since it is irrelevant
measles_year$filter <- NULL 

# adding a single "week" column
# changing the data type from chr to date
measles_week <- measles_week |> mutate(week = week_start) |> select(week, everything())
measles_week$week <- as.Date(measles_week$week)


# cleaning the measles_by_state data
# we are choosing the max except for 250+
temp <- rep(0, length(measles_state$cases_range))
temp[measles_state$cases_range == "0"] <- 0
temp[measles_state$cases_range == "1-9"] <- 9
temp[measles_state$cases_range == "10-49"] <- 49
temp[measles_state$cases_range == "50-99"] <- 99
temp[measles_state$cases_range == "250+"] <- 250

measles_state <- measles_state |> mutate(case_num = temp)
remove(temp)

# cleaning the vaccine_by_state data
vaccine_state$estimate_pct <- as.numeric(gsub("%","", vaccine_state$estimate_pct))

```

Time series plot for measles cases

```{r}
measles_week_plot <- ggplot(measles_week, aes(x = week, y = cases)) + 
  geom_line() + 
  labs(
    x = "Date",
    y = "Cases",
    title = "Measles Cases Since 2023",
    caption = "Source: CDC"
  ) + 
  geom_vline(xintercept = as.Date("2025-01-20"), color = 'red', linewidth = 1.2) + 
  theme_bw()
measles_week_plot
```


## Preparation for geospatial data

```{r}

us_states <- map_data('state')

measles_state <- measles_state |> mutate(region = tolower(geography))
measles_map <- left_join(us_states, measles_state, by = 'region', relationship = "many-to-many")


vaccine_state <- vaccine_state |> mutate(region = tolower(geography))


vacc_map <- left_join(us_states, vaccine_state, by = "region", relationship = "many-to-many")

```



## Visualizting infection rates and vaccination by states

```{r}
plot_measles <- plot_vaccine <- ggplot(measles_map |> filter(year == "2025"), aes(x = long, y = lat, group = group, fill = case_num)) +
  geom_polygon(color = "darkgrey") +
  coord_fixed(1.3) +
  scale_fill_gradient(name = "Measles Cases", high = "red", low = "white") +
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Measles Cases by State",
    caption = "Source: CDC"
    ) + 
  theme_minimal()


plot_vaccine <- ggplot(vacc_map, aes(x = long, y = lat, group = group, fill = estimate_pct)) +
  geom_polygon(color = "darkgrey") +
  coord_fixed(1.3) +
  scale_fill_gradient(name = "Vaccination Rate (%)", high = "green", low = "red") +
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Vaccination Rates by State",
    caption = "Source: CDC"
    ) + 
  theme_minimal()

ggsave("cases_vs_vaccine.png", plot = plot_measles + plot_vaccine, width = 18, height = 8, dpi = 300)
```

