---
title: "Aggregate Predictors of Mental Health Outcomes"
output: html_notebook
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidycensus)
library(tidyverse)
library(ggplot2)
library(readr)
```

## Main guiding questions

1. At the province/state level, are mental health outcomes (e.g. prevalence of poor mental health, hospitalizations for mental illness) more strongly associated with demographic composition (e.g. % youth, % female, % older adults) or socioeconomic indicators (median income, % with college degree, unemployment rate)?

2. Do environmental factors (e.g. heat waves, cold spells) appear correlated with changes in population-level mental health outcomes?

3. How do these associations vary over time (e.g. pre/post pandemic or across several years)?



## Data Source

- [CDC Data Explorer](https://ephtracking.cdc.gov/DataExplorer/)
  - Source 1: Age-adjusted prevalence of frequent mental distress among adults (model-based with confidence intervals)
  - Source 2: Age-adjusted Prevalence of Depression Among Adults (model-based with confidence intervals)
  - The methodology for the estimation can be found [here](https://www.cdc.gov/places/methodology/index.html)
- [US Census Data](https://data.census.gov/), integrated using the `tidycensus` [R package](https://walker-data.com/tidycensus/).


## Loading the data

```{r}
depression_og <- read_csv("data/cdc_depression_adjusted.csv")
distress_og <- read_csv("data/cdc_mental_distress_adjusted.csv")
```

## Cleaning

```{r depression cleaning}
depression <- depression_og |> 
  rename(
    year = Year,
    LCI = "Confidence Interval Low",
    UCI = "Confidence Interval High", 
    prevalence_distress = Value) |>
  filter(year %in% c(2019, 2020, 2021)) |> 
  select(State, County, year, prevalence_distress, LCI, UCI)
```


```{r distress cleaning}
distress <- distress_og |> 
  rename(
    year = Year,
    LCI = "Confidence Interval Low",
    UCI = "Confidence Interval High", 
    prevalence_distress = Value) |>
  filter(year %in% c(2019, 2020, 2021)) |> 
  select(State, County, year, prevalence_distress, LCI, UCI)
```

