---
title: "Assignment 01"
author: "Your Name and UID"
date: "Due: 11:59 PM, September 8, 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Submission Instruction:

- Submit: a single HTML document
- Show all code and output


As a reminder, the cheatsheet for R Markdown can be found in `Help` --> `Cheatsheets`.

## Exercise 1

Set a seed. Sample 10 times from a $\mathcal{N}(0,1)$ distribution.  
```{r}
# Your code
```

1.  Compute the mean and standard deviation of your sample. 
```{r}
# Your code, same below and in the future.
```

2.  Repeat the experiment with samples of size 100 and 1000 and 10000 and 100000. 

3.  Using the `plot` function, plot the estimated mean (on the y-axis) versus the sample size (on the x-axis).  

4.  Create another plot for the estimated standard deviation.


## Exercise 2

1. Create a list/vector `v` with the following entries:
  - the character ”H” repeated 5 times,
  - the character ”e” repeated 5 times,
  - the character ”l” repeated 5 times,
  - the character ”o” repeated 5 times,
  - the character ”W” repeated 5 times,
  - the character ”o” repeated 5 times,
  - the character ”r” repeated 5 times,
  - the character ”l” repeated 5 times,
  - the character ”d” repeated 5 times.

```{r}
# Your code
```

 
2.  Extract the sublist/subvector composed of every fifth entry of `v` (starting with the second entry, so you would extract the 2nd, 7th, 12th,..., entries). Print the list. 



