---
title: "Weather Data"
author: "Fushuai Jiang"
output: 
  html_document:
    toc: TRUE
---

Notes for real-time visualizing weather data for the Four-Corner neighborhood MD 20901. 


```{r setup, include = FALSE}

# Global options for chunks
# echo set to TRUE so all codes are displayed unless specified otherwise
knitr::opts_chunk$set(echo = TRUE)

library(httr)
library(jsonlite)
library(lubridate)
library(tidyverse)
library(ggplot2)
library(ggthemes)
```

# Loading the weather data

First, we load the data from Open-Meteo from <https://api.open-meteo.com/v1/forecast>.The data is open-sourced so no api key is needed. 

```{r data loading}
# Define URL and parameters
url <- "https://api.open-meteo.com/v1/forecast"

params <- list(
  latitude = 39.02,
  longitude = 77.01,
  hourly = "temperature_2m,relative_humidity_2m,rain,visibility,wind_speed_10m", # we select variables including temperature, humidity, rainfall, visibility, and windspeed. Make sure there is no space between the variable names
  current = "temperature_2m,relative_humidity_2m,rain",
  timezone = "America/New_York",
  past_days = 7,
  wind_speed_unit = "mph",
  temperature_unit = "fahrenheit",
  precipitation_unit = "inch"
)

response <- GET(url, query = params)

# Parse JSON content
weather_data <- fromJSON(content(response, "text"))

# Convert hourly data into a dataframe
hourly_data <- weather_data$hourly
df <- as.data.frame(hourly_data)
df$time <- ymd_hm(df$time, tz = "America/New_York")

# Convert current data into a dataframe
df_current <- as.data.frame(weather_data$current)
df_current$time <- ymd_hm(df_current$time, tz = "America/New_York")
```

We can also save the dataset locally, with the date including in its name. 

```{r}

today_str <- format(Sys.Date(), "%Y%m%d")
write.csv(df, file = paste0("weather_data_prediction", today_str,".csv"), row.names = FALSE)
write.csv(df_current, file = paste0("weather_data_today", today_str,".csv"), row.names = FALSE)
message("Data saved locally.")

```


A quick look at the weather_data.

```{r data quick view}
str(df)
slice_sample(df, n = 20)
```



# Temperature visualization

```{r temperature viz}


# Plot the temperature
g_temp <- ggplot(df, aes(x = time, y = temperature_2m)) + 
  geom_line(linewidth=0.5, linetype = 1, color = "steelblue") + 
  geom_smooth(color = "darkgreen") + 
  geom_point(data = df_current, shape = 17, size = 5, color = "red") + 
  geom_text(data = df_current, 
            aes(
              x = time, 
              y = temperature_2m, 
              label = paste0("Now: ", round(temperature_2m, 1), "°F")
              ),
            vjust = -1, color = "red", fontface = "bold") +
  scale_x_datetime(
    date_breaks = "1 day"
  ) +
  labs(
    title = "Temperature in Four Corners (+/- 7 days)",
    x = "Time and Date", 
    y = "Temperature (Fahrenheit)",
    caption = "Source: Open-Meteo weather data"
    ) + 
  theme_bw() + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

g_temp

```
# Humidinity visualization

```{r humidity viz}

# Plot the temperature
g_humid <- ggplot(df, aes(x = time, y = relative_humidity_2m)) + 
  geom_line(linewidth=0.5, linetype = 1, color = "steelblue") + 
  geom_smooth(color = "darkgreen") + 
  geom_point(data = df_current, shape = 17, size = 5, color = "red") + 
  geom_text(data = df_current, 
            aes(x = time, y = relative_humidity_2m, 
                label = paste0("Now: ", round(relative_humidity_2m, 1), "%")),
            vjust = -1, color = "red", fontface = "bold") +
  labs(
    title = "Relative Humidity in Four Corners (+/- 7 days)",
    x = "Time and Date", 
    y = "Temperature (Fahrenheit)",
    caption = "Source: Open-Meteo weather data"
    ) + 
  theme_bw()
g_humid

```

# Overlay temperature and humidity


```{r}
# Compute scaling factor to align the two y-axes visually
# E.g., if temp ~ 70°F and humidity ~ 50%, scale humidity to same range
# range_temp <- range(df$temperature_2m, na.rm = TRUE)
# range_humidity <- range(df$relative_humidity_2m, na.rm = TRUE)

range_temp <- c(60,100)
range_humidity <- range(0,100)

scale_factor <- diff(range_temp) / diff(range_humidity)
# scale_factor <- 50/100

# Build plot
g_temp_vs_humid <- ggplot(df, aes(x = time)) +
  # Temperature line
  geom_line(aes(y = temperature_2m, color = "temp_color"), size = 0.8) +
  # Humidity line, scaled to match temp range
  geom_line(aes(y = relative_humidity_2m * scale_factor + min(range_temp), color = "humid_color"), size = 0.8, linetype = 4) +
  
  # Primary y-axis for temperature
  scale_y_continuous(
    name = "Temperature (F)",
    limits = range_temp,
    # Secondary y-axis: rescale back to humidity
    sec.axis = sec_axis(~ (. - (min(range_temp)))/scale_factor, 
                        name = "Humidity (%)", 
                        breaks = seq(0,100,20)
                        )
  ) +
  scale_x_datetime(
    date_breaks = "1 day"
  ) + 
  # Colors
  scale_color_manual(values = c("temp_color" = "firebrick", 
                                "humid_color" = "darkgreen")) +
  
  labs(
    title = "Temperature and Humidity in Four Corners MD 20901 (+/- Days)",
    x = "Time", 
    color = "", 
    caption = "Source: Open-Meteo https://api.open-meteo.com") +
  theme_minimal() +
  theme(
    axis.title.y.left = element_text(color = "firebrick"),
    axis.text.y.left = element_text(color = "firebrick"),
    axis.title.y.right = element_text(color = "darkgreen"),
    axis.text.y.right = element_text(color = "darkgreen"),
    legend.position = "top",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

g_temp_vs_humid
```
We see a potential negative correlation between temperature and relative humidity. A further look into the correlation factor confirms this. 

```{r}
corr_temp_humid <- cor(df$temperature_2m,df$relative_humidity_2m)
cat(paste0("The correlation factor between temperature and the relative humidity is ", round(corr_temp_humid, 2), ".\n"))
```

Here, we also showcase the in-line code mode of Rmarkdown, where R code can be embedded. For instance, the correlation computed above was `r cor(df$temperature_2m,df$relative_humidity_2m)`. 



# Rain and Wind

We now take a look at the wind data. 

```{r}

g_storm <- ggplot(df, aes(x = time)) + 
  geom_col(aes(x = time, y = wind_speed_10m), fill = "blue") + 
  scale_x_datetime(
    date_breaks = "1 day"
  ) + 
  labs(
    y = "Wind Speed (mph)",
    x = "Time & Date",
    caption = "Source: Open-Meteo",
    title = "Wind Speed in Four Corners MD 20901 +/-7 days"
  )+ 
  theme_solarized() + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

g_storm

```
After examing the correlation between wind speed and other variables, we see that wind speed is the most (negatively) correlated with the relative humidity with a correlation factor of `r cor(df$wind_speed_10m,df$relative_humidity_2m)`, so let's overlay the plot.

```{r, include = FALSE}

cor(df$wind_speed_10m, df$temperature_2m)
cor(df$wind_speed_10m,df$rain)
cor(df$wind_speed_10m,df$relative_humidity_2m)
```


```{r}
# Compute scaling factor to align the two y-axes visually
# E.g., if temp ~ 70°F and humidity ~ 50%, scale humidity to same range
# range_temp <- range(df$temperature_2m, na.rm = TRUE)
# range_humidity <- range(df$relative_humidity_2m, na.rm = TRUE)

range_wind <- c(0,20)
range_humidity <- range(0,100)

scale_factor <- diff(range_wind) / diff(range_humidity)
# scale_factor <- 50/100

# Build plot
g_wind_vs_humid <- ggplot(df, aes(x = time)) +
  # Temperature line
  geom_line(aes(y = wind_speed_10m, color = "Wind Speed"), size = 0.8) +
  # Humidity line, scaled to match temp range
  geom_line(aes(y = relative_humidity_2m * scale_factor + min(range_wind), color = "Humidity"), size = 0.8, linetype = 4) +
  
  # Primary y-axis for temperature
  scale_y_continuous(
    name = "Wind Speed (mph)",
    limits = range_wind,
    # Secondary y-axis: rescale back to humidity
    sec.axis = sec_axis(~ (. - (min(range_wind)))/scale_factor, 
                        name = "Humidity (%)", 
                        breaks = seq(0,100,20)
                        )
  ) +
  scale_x_datetime(
    date_breaks = "1 day"
  ) + 
  scale_color_manual(
    values = c(
      "Wind Speed" = "blue",
      "Humidity" = "darkgreen"
    )
  ) + 
  labs(
    title = "Wind Speed and Humidity in Four Corners MD 20901 (+/- 7 Days)",
    x = "Date/Time", 
    color = "", 
    caption = "Source: Open-Meteo https://api.open-meteo.com") +
  theme_minimal() +
  theme(
    axis.title.y.left = element_text(color = "blue"),
    axis.text.y.left = element_text(color = "blue"),
    axis.title.y.right = element_text(color = "darkgreen"),
    axis.text.y.right = element_text(color = "darkgreen"),
    legend.position = "top",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

g_wind_vs_humid
```


# Data transformation

Suppose we want to visualize the temperature differently. Instead of a curve showing daily oscillation, we want a bar for each day showing the max and the min. In this case, we need to do a little data manipulation, first grouping the data by date and compute the daily min/max. 

```{r}

df_tempbyday <- df |>
  # first suppress all the time
  mutate(date  = as.Date(time)) |>
  # Then we want to group the data according to each day
  # This is mostly in preparation for the next step
  group_by(date) |>
  summarize(
    temp_min = min(temperature_2m),
    temp_max = max(temperature_2m)
  )
str(df_tempbyday)
```


Now we are ready to plot. We will make use of `geom_linerange()`. 

```{r}
g_tempbyday <- ggplot(df_tempbyday, aes(x = date)) + 
  geom_linerange(aes(
    ymin = temp_min,
    ymax = temp_max,
  ), color = "darkgreen", 
    size = 2) + 
  geom_point(aes(y = 0.5*(temp_min + temp_max)), color = "red", size = 3) + 
  labs(
    title = "Temperature by Day",
    x = "Date",
    y = "Temperature (F)",
    caption = ""
  ) + 
  theme_tufte()

g_tempbyday
```



# Compare with historical data

Note that Open-Meteo keeps track of past weather data dating back to 1940 <https://archive-api.open-meteo.com/v1/archive>. However, there is a 5-day delay in the data. This loops take a while to run, so we can use messages to keep us assured. 

```{r}

all_data <- list()
# df_temp_hist <- data.frame(year = NA, time = NA, temperature_2m = NA)
  
  for (year in 1950:2025) {
    # Build start and end dates
    # need to reformat the system date to get rid of the year
    start_date <- paste0(year, "-", format(Sys.Date()-13, "%m-%d"))
    end_date <- paste0(year, "-", format(Sys.Date()-7, "%m-%d"))
    
    # Build params
    params_temp_hist <- list(
      latitude = 39.02,
      longitude = 77.01,
      start_date = start_date,
      end_date = end_date,
      hourly = "temperature_2m",
      timezone = "America/New_York",
      temperature_unit = "fahrenheit"
    )
    
    # Make API call to historical archive endpoint
    url_hist <- "https://archive-api.open-meteo.com/v1/archive"
    response_hist <- GET(url_hist, query = params_temp_hist)
    
    # Parse JSON content
    weather_data_hist <- fromJSON(content(response_hist, "text", encoding = "UTF-8"))
    
    if (!is.null(weather_data_hist$hourly)) {
      df_year <- as.data.frame(weather_data_hist$hourly)
      df_year$time <- ymd_hm(df_year$time, tz = "America/New_York")
      df_year$year <- year
      # bind_rows(df_year, as.data.frame(df_temp_hist))
      all_data[[as.character(year)]] <- df_year
      message("Retrieved data for year: ", year)
    } else {
      warning("No data returned for year: ", year)
    }
    
    # Optional: polite pause to avoid rate limits
    Sys.sleep(0.3)
  }
  
  # Combine into one dataframe
  df_temp_hist <- bind_rows(all_data)
  
 

```

First we consider a messy line plot of temperature, each line corresponds to a single year. There are some preliminary steps to format the data. 

1.  We want to format the year as factors (categorical variable)
2.  We want to eliminate the year in front of the time so that ggplot will think of all the years as being in the same timeline

```{r}
df_temp_hist$time <- format(df_temp_hist$time, format = "%m-%d %H:%M:%S", tz = "America/New_York")

ggplot(df_temp_hist) + 
  geom_line(aes(x = time, y = temperature_2m, group = as.factor(year), color = as.factor(year)))

```

Of course, this is way too messy. Let's select a subset of these with 10 years gap. 

```{r}

df_temp_hist$time_noyear <- as.POSIXct(
  paste0("2000-", df_temp_hist$time),  # add dummy year
  format = "%Y-%m-%d %H:%M:%S",
  tz = "America/New_York"
)


```


```{r}


ggplot(subset(df_temp_hist, year %in% seq(1955,2025,10))) + 
  geom_line(aes(x = time_noyear, y = temperature_2m, color = as.factor(year)), size=0.4, alpha=0.7) +
  scale_x_datetime(date_breaks = "1 day", date_labels = "%m-%d") +
  labs(
    title = "Temperature over July 1–8 across decades",
    x = "Date & Time (July 1–8)",
    y = "Temperature (F)",
    color = "Year"
  ) +
  theme_tufte() +
  theme(
    axis.text.x = element_text(angle=45, vjust=0.8, hjust=1)
  )

```
While the plot above doesn't really tell a definite story, we can do one step further and compute the historical averages of July 1 - July 8 and see if there is any trend. 

```{r}
df_temp_hist_avg <- df_temp_hist |> group_by(year) |> summarize(mean(temperature_2m))
df_temp_hist_avg <- df_temp_hist_avg |> rename(temp_avg = 'mean(temperature_2m)')
```

After obtaining the historical averages, let's look at a line plot.

```{r}
ggplot(df_temp_hist_avg, aes(x = year, y = temp_avg)) + 
  geom_line() + 
  geom_point(color = "red", size = 2.5) + 
  geom_smooth(method = "lm")
```
The average seems to be linearly decreasing from the graph. However, let's include the the max temp and the min temp to see what happens. First we collect the min, max, and median. 

```{r}
df_temp_hist_stat <- df_temp_hist |> 
  group_by(year) |> 
  summarize(
    temp_max = max(temperature_2m),
    temp_min = min(temperature_2m),
    temp_med = median(temperature_2m)
  )
```

Then we plot.

```{r}
ggplot(df_temp_hist_stat, aes(x = year)) + 
  geom_line(aes(y = temp_max, color = "Max Temperature"), ) + 
  geom_line(aes(y = temp_min, color = "Min Temperature")) + 
  geom_line(aes(y = temp_med, color = "Median Temperature")) +
  scale_color_manual(
    values = c("Max Temperature" = "red",
    "Min Temperature" = "blue",
    "Median Temperature" = "darkgreen"), 
    name = NULL
  ) + 
  geom_smooth(method = "lm", aes(y = temp_max), color = "darkred", alpha = 0.6) + 
  geom_smooth(method = "lm", aes(y = temp_min), color = "darkblue", alpha = 0.6) + 
  geom_smooth(method = "lm", aes(y = temp_med), color = "green", alpha = 0.6) + 
  labs(
    title = "Historical Averages",
    x = "Year", 
    y = "Temperature",
    caption = "Source: Open Meteo"
  ) + 
  theme_tufte()
```
Let's try bar plot instead and overlay a smoothing model over, where the smoothing model is over all historical data. 

```{r}
ggplot(df_temp_hist_stat, aes(x = year)) + 
  geom_linerange(aes(
    ymin = temp_min,
    ymax = temp_max,
  ), color = "darkgreen", 
    size = 1.3, alpha = 0.6) + 
  geom_point(aes(y = temp_med), color = "red", size = 2.5, shape = 12) + 
  geom_smooth(data = df_temp_hist, method = "lm", aes(x = year, y = temperature_2m)) + 
  labs(
    title = "Historical Averages",
    x = "Year",
    y = "Temperature (F)",
    caption = "Source: Open Meteo Archive"
  ) + 
  theme_tufte()
```

