---
title: 'Dataframe manipulation'
author: "Data350"
date: "2024-09-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Warm up: *cars* dataset

Take a look at the data set (and plot it for fun if you like). The command *cars\$column_name* will extract the single column with the name "column_name".

1.  Write two lines of code to summarize each individual column.

```{r}
summary(cars$dist)
summary(cars$speed)
```

2.  Create a list called slow_ones (of Booleans) indexing the cars with speed less than 10. Now replace the speed of every slow cars by "Slow"

```{r}
slow_ones <- cars$speed < 10 # indicates which row should be slow 
# think of slow_ones as a index vector telling us which rows to pick
# cars$speed[slow_ones] # will pick out the slow ones according to the index
cars$speed[slow_ones] <- "Slow"
cars # note that the first column is now character
```

3.  Do the same thing but pick out cars with speed in $[10,20)$, replace their speed by "Alright"; the ones $\geq 20$ are called "Tracy Chapman"

```{r}
cars$speed[cars$speed >= 10 & cars$speed < 20] <- "Alright"
cars$speed[cars$speed > 20] <- "Tracy Chapman"
cars ## Doesn't work!!!!!!
```

We have truly messed up the dataset, so let's use *rm(cars)* to get rid of the evidence.

```{r}
rm(cars)
sp <- numeric(50) # create a list of 50 zeros
sp[cars$speed < 10] <- "Slow"
sp[cars$speed >= 10 & cars$speed < 20] <- "Alright"
sp[cars$speed >= 20] <- "Tracy Chapman forever"
# sp is a list of characters corresponding to the labels
cars$speed <- sp
cars
```

# Data transformation in tidyverse

Based on Chapter 3 of Wickham, Hadley, Mine Çetinkaya-Rundel, and Garrett Grolemund *R for data science*.

Load the necessary library. The dataset is called "flights".

```{r}
library(tidyverse)
library(nycflights13)
```

1.  Read about the dataset *flight*, its dimension, the data type, etc.

## filter()

2.  The *filter()* command allows us to subset observations based on their values. What does *filter(flights, month == 12 | day == 14)* do?

```{r}
filter(flights, month == 12 | day == 14) # | is the logical OR, INCLUSIVE!!!
filter(flights, month == 12, day == 14) # comma is the logical AND, the same as
filter(flights, month == 12 & day == 14)
```

How to view all of the flights that were delayed on arrival on the date June 1?

```{r}

# filter(flights, month == 6, day == 1, arr_time > sched_arr_time) 
# this one actually doesn't quite work, because some flights arrive after midnight
filter(flights, month == 6, day == 1, arr_delay > 0)
# note that the dimension are different, so better to insert rm.na = TRUE
```

3.  What does the following command do? Read about *table()*.

```{r}
NovDec <- filter(flights, month %in% c(11,12)) # filter out flights in November or December
table(NovDec$month)
# table() will count the frequencies
```

To keep the "NA" data, add *is.na()* into the filter

```{r}
NovDec <- filter(flights, is.na(month), month ==11 | month == 12) # this keeps the NA data
table(NovDec$month)
```

## arrange()

You can re-order rows for easier viewing by using *arrange()*.

1.  Arrange flights according to arrival delay and then departure delay.

```{r}
arrange(flights, dep_delay, arr_delay)
```

2.  The default is ascending, but you can use *desc()* to change the order

```{r}
arrange(flights, desc(dep_delay), arr_delay)
```


## select()

The *select()* command allows us to pick variables by name. It allows us to zoom in on useful subsets of the variables efficiently. allows us to zoom in on useful subsets of the variables efficiently.

1.  Try

```{r}
# created a new dataframe by selecting the columns dep_delay, arr_dealy, and carrier from flights
CarrierDelay <- select(flights, dep_delay, arr_delay, carrier)
```

What does the following code do?

```{r}
# remove all columns from year to sched_arr_time
select(flights, -c(year:sched_arr_time))
```

2.  *select()* has some useful helper functions:

-   starts_with( “abc” )
-   ends_with( “xyz” )
-   contains( “ijk” )
-   num_range(“x”,1:3) matches x1, x2, x3

Select the columns that have everything to do with departure and arrival time.

```{r}
select(flights, contains("arr") | contains("dep"), -c("carrier"))
```


3.  you can add *everything()* in *select()* to move the columns you select to the front. Try it!

4.  You can use rename() to rename a column.

# The pipe \|\>

Consider how you will accomplish the following sequence of tasks:

1.  Filter the flights so they all fly into the DC area airport: IAD, BWI, DCA

```{r}
flights1 <- filter(flights, dest %in% c("IAD", "BWI", "DCA"))
```

2.  Create a new column called speed (mph) = distance / (air time/60)

```{r}
flights2 <- mutate(flights1, speed = distance / (air_time/ 60))
```

3.  Pick out the columns including dates, departure time, carrier, flight number, speed

```{r}
flights3 <- select(flights2, year:day, dep_time, carrier, flight, speed)
```

4.  Arrange those rows by speed in descending order.

```{r}
arrange(flights3, desc(speed))
```

If you want to code everything by brute force:

```{r}
# impossible to read!

arrange(
  select(
  mutate(
  filter(flights, dest %in% c("DWI","IAD", "DCA")),
  speed = distance / (air_time/60)
),
  year:day, dep_time, carrier, flight, speed
),
  desc(speed)
)

```

The pipe \|\> simplifies the procedure. The general syntax is *DATA \|\> FUNCTION_1(...) \|\> FUNCTION_2(...) \|\> ...*

```{r}
flights |> 
  filter(dest %in% c("BWI", "IAD", "DCA")) |> # apply filter
  mutate(speed = distance/(air_time/60)) |> # create a new varible
  select(year:day, dep_time, carrier, flight, speed) |> # selection columns
  arrange(desc(speed)) # sorting
```

### Exercise 1

```{r}

```

Do the following sequence of tasks.

1.  Look up "with" command, then change the "data" column by ISO format using "flights\$date \<- with(flights, ISOdate(year, month, day))"

```{r}
flights$date <- with(flights, ISOdate(year, month, day))
```

2.  group the flights by date
3.  summarize the average delay for each day, make sure to remove missing data by inserting "na.rm = TRUE"
4.  Create a line plot of daily delay, with x label "Date" and y label "Average Departure Delay"

```{r}
flights$date <- with(flights, 	ISOdate(year, month, 	day))

flights |>
  group_by(date) |> # grouped by date
  summarize(delay = mean(dep_delay,na.rm = TRUE)) |>
  ggplot(
    mapping = aes(x = date, y = delay)
  )  +        # I have created a graph object
  geom_line() + 
  labs(x = "Date", y = "Average Departure Delay")
```

### Exercise 2

Let's create a **bar plot** counting the average departure delay for each carrier.

# Visualizing statistics

```{r}
library(gapminder)
```

Do the following.

1.  In the gapminder dataset, only consider the data in the years 1952, 1972, and 1992.

2.  Create a point plot of life expectancy vs GDP per capita, and Create a regression curve for each year.

-   Use color to different the years, and use size to describe population size.
-   Play around with different regression models to find the best one (to tell your story).
-   You may want to rescale the axis to make things look nicer. Type in "scale" in the Help window and take a look.

4.  Fix the legends, pick themes, etc.

5.  Write everything using the pipe (if you haven't already).
