---
title: "Quiz 3 prep"
author: "Fushuai Jiang"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(ggplot2)
library(dplyr)
library(socviz)
```






All of the exercises will be using the sub-sampled dataset `df`. 

```{r}
df <- gss_sm |> 
  select(age, wtssall, childs, sex, income16, bigregion, partyid, zodiac) |> 
  slice_sample(n = 2000)
```



1. Write a minimal shiny app that allows the user to pick from `c("age", "wtssall", "childs")` and output the five-number statistics using `summary()` and `verbatimTextOutput`.


```{r}
library(shiny)

ui <- fluidPage(

)

server <- function(input, output, session) {

}

shinyApp(ui, server)
```


2. Add an action button to the app in 1 with text "Five Numbers", such that the output will only display/update after you click the button. 


```{r}

library(shiny)

ui <- fluidPage()

server <- function(input, output, session) {}

shinyApp(ui, server)
```


3. Write a shiny app that allow the user to choose from `c("age", "wtssall", "childs")` and display a density plot of that variable using `geom_density`. 

```{r}
library(shiny)

ui <- fluidPage()

server <- function(input, output, session) {}


shinyApp(ui, server)
```


4. Build on top of the app in part 3 and accomplish the following:

  - Add an action button with text "Visualize" so the plot only displays after clicking the button
  - Facet the plot by zodiac sign
  
```{r}
library(shiny)

ui <- fluidPage()

server <- function(input, output, session) {}

shinyApp(ui, server)
```


5. (The quiz won't be this hard) Build on top of part 4, so that the user can choose faceted variables from `c("zodiac", "partyid", "bigregion")`


```{r}
library(shiny)

ui <- fluidPage()

server <- function(input, output, session) {}

shinyApp(ui, server)
```

