---
title: '20251111'
author: "Data350"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Last time (20251106):

 - basic structure of shiny app
 - type "shinyapp" in an R script and pick the snippet to get started
 - can also embed a shiny app into an HTML (New File --> Markdown --> Shiny Document)


### Inside the UI
 - Input syntax: `{method}Input("inputID", ...)`
 - Output syntax: `{method}Output("outputID", ...)`



### Example input functions
 - `numericInput("id", "Label", value = , min =, max = )`
 - `sliderInput("id", "Label", value = , min =, max = )`
 - `selectInput()` to choose from a list of preset options
 - `radioButtons()`, similar function (`choiceNames`/`choiceValues`) arguments allow you to display options other than plain text
 - `checkboxGroupInput()` allows for multiple options (as opposed to `chekboxInput()`)


### Inside the server

- Calling the input: `input$inputID`
- Calling the output: `output$outputID`
- Render syntax: `render{Type}({...})`


###  List of render functions
 - `renderText` <--> `textOutput`
 - `renderPrint` <--> `verbatimTextOutput`
 - `renderTable` <--> `tableOutput`
 - `renderDataTable` <--> `dataTableOutput`
 - `renderPlot` <--> `plotOutput`
 
 
 

## Today

### Warm up exercise


Build a Shiny app to visualize `mpg`:

  - input: choice of year using radio buttons
  - output: scatterplot with `displ` against `hwy`

 
 

### Basic reactivity

Read [Shiny Section 3.3](https://mastering-shiny.org/basic-reactivity.html#reactive-programming). 


## Trivial Example

```{r}
library(shiny)

ui <- fluidPage(
  textInput("name", "What's your name?"),
  textOutput("greeting")
)

server <- function(input, output, session) {
  output$greeting <- renderText({
    paste0("Hello ", input$name, "!")
  })
}


shinyApp(ui,server, session)
```





## Two-sample t-test




```{r, out.width="65%"}
library(shiny)
library(ggplot2)

## Customized functions taken from 3.4.1

freqpoly <- function(x1, x2, binwidth = 0.1, xlim = c(-3, 3)) {
  df <- data.frame(
    x = c(x1, x2),
    g = c(rep("x1", length(x1)), rep("x2", length(x2)))
  )

  ggplot(df, aes(x, colour = g)) +
    geom_freqpoly(binwidth = binwidth, size = 1) +
    coord_cartesian(xlim = xlim)
}

t_test <- function(x1, x2) {
  test <- t.test(x1, x2)

  # use sprintf() to format t.test() results compactly
  sprintf(
    "p value: %0.3f\n[%0.2f, %0.2f]",
    test$p.value, test$conf.int[1], test$conf.int[2]
  )
}




## UI

ui <- fluidPage(
  fluidRow(
    column(4,
      "Distribution 1",
      numericInput("n1", label = "n", value = 1000, min = 1),
      numericInput("mean1", label = "µ", value = 0, step = 0.1),
      numericInput("sd1", label = "σ", value = 0.5, min = 0.1, step = 0.1)
    ),
    column(4,
      "Distribution 2",
      numericInput("n2", label = "n", value = 1000, min = 1),
      numericInput("mean2", label = "µ", value = 0, step = 0.1),
      numericInput("sd2", label = "σ", value = 0.5, min = 0.1, step = 0.1)
    ),
    column(4,
      "Frequency polygon",
      numericInput("binwidth", label = "Bin width", value = 0.1, step = 0.1),
      sliderInput("range", label = "range", value = c(-3, 3), min = -5, max = 5)
    )
  ),
  fluidRow(
    column(9, plotOutput("hist")),
    column(3, verbatimTextOutput("ttest"))
  )
)

server <- function(input, output, session) {
  output$hist <- renderPlot({
    x1 <- rnorm(input$n1, input$mean1, input$sd1)
    x2 <- rnorm(input$n2, input$mean2, input$sd2)

    freqpoly(x1, x2, binwidth = input$binwidth, xlim = input$range)
  }, res = 96)

  output$ttest <- renderText({
    x1 <- rnorm(input$n1, input$mean1, input$sd1)
    x2 <- rnorm(input$n2, input$mean2, input$sd2)

    t_test(x1, x2)
  })
}


shinyApp(ui, server, session)



```




## Modify the `mpg` visualizer with an action button








